---
title: "Overview"
---

```{r, include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

Over the course of this 1-day training class you will learn to run
water use scenarios with the Wood River Valley (WRV) groundwater-flow model.

## Tentative Schedule

| Time         | Description | Instructor |
|:------------:| -------------------------------------------------------------------------------- | --------------- |
| 8:00--8:15   | Welcome and introductions                                                        |                 |
| 8:15--9:15   | Overview of WRV groundwater-flow-model publication, archive, and repository      | J. Fisher, USGS |
| 9:15--9:35   | Break                                                                            |                 |
| 9:30--10:30  | Introduction to the R-programming language                                       | J. Fisher, USGS |
| 10:45--11:45 | Running and analysing the archived [calibrated model](calib-model.html)          | J. Fisher, USGS |
| 11:45--12:45 | Lunch                                                                            |                 |
| 12:45--1:30  | Running and analyzing the calibrated model                                       | J. Sukow, IDWR  |
| 1:30--2:00   | Running scenarios using the fully populated model, overview of example scenarios | J. Sukow, IDWR  |
| 2:00--2:15   | Break                                                                            |                 |
| 2:15--3:00   | Scenario 1: an example simulation of managed recharge                            | J. Sukow, IDWR  |
| 3:00--4:30   | [Scenario 2](scenario-2.html): an example simulation of idling irrigated lands   | J. Fisher, USGS |

## Hardware

Requires an internet connection and about 16 gigabytes of memory.

## Software

If [R](https://www.r-project.org/) (version 3.3 or greater) is not already installed on your computer,
download and install the latest binary distribution from the
**C**omprehensive **R** **A**rchive **N**etwork ([CRAN](https://cran.r-project.org/package=inlmisc)).
Installation instructions for installing R are located at
https://jfisher-usgs.github.io/r-workshop/software.html#r.
R should be accessible from the start menu, desktop shortcut, or command shell.
Assuming you installed R on a 64-bit operating system (OS), two versions of R are made available,
that is, a 32-bit and 64-bit version.
The 64-bit version of R (R x64) is preferred.
An introduction to the R-programming language is provided at
https://jfisher-usgs.github.io/r-workshop.

Startup an R session.
Extend the capabilities of R by installing the R-package **wrv**.
You can install the stable version of **wrv** from the
**G**eological Survey **R** **A**rchive **N**etwork ([GRAN](https://owi.usgs.gov/R/gran.html)),
and its dependencies from CRAN,
by pasting the following commands into your R console:

```{r}
repos <- c(GRAN = "http://owi.usgs.gov/R", CRAN = "https://cloud.r-project.org/")
update.packages(ask = FALSE, repos = repos)
install.packages("wrv", repos = repos, dependencies = TRUE)
```

You will only need to install R packages once on your computer.
To ensure that everything got installed properly,
load the **wrv** package in your current R session using

```{r}
library("wrv")
```

## Files

The groundwater-flow model is described in the associated
[model report](http://dx.doi.org/10.3133/sir20165080 "USGS Scientific Investigations Report") and
[model archive](http://dx.doi.org/10.5066/F7C827DT "USGS Data Release").
Processing programs and unprocessed data sets for the model are bundled together in an R package named **wrv**---the
file repository for this package is located on [GitHub](https://github.com/USGS-R/wrv).
Included in the **wrv** package is [MODFLOW-USG](http://water.usgs.gov/ogw/mfusg/ "MODFLOW-USG") version 1.3,
a U.S. Geological Survey (USGS) computer code that solves the groundwater-flow equation.

Files used in this training class were placed online in the
[wrv-training repository](https://github.com/jfisher-usgs/wrv-training).
Create a copy of the repository on your computer using R.
To do so, first specify where you want to place the repository in your local file system.
Type the following commands substituting in the desired path;
for example, I changed the path from `"<path/to/repository>"` to `"I:/Software/wrv-training"`
making sure to use the forward slash (/) to separate folder names:

```{r}
path <- "<path/to/repository>"
dir.create(path, recursive = TRUE)
```

Next, copy the remote repository to your computer.

```{r}
git2r::clone("https://github.com/jfisher-usgs/wrv-training.git", path)
```

All subsequent R instructions require that the *working directory* be set to the local wrv-training repository every time you start an R session.
A working directory allows us to use relative paths instead of full absolute paths in our R scripts.
An absolute path points to the same location in a file system, regardless of the current working directory.
By contrast, a relative path starts from some given working directory, avoiding the need to provide the full absolute path.
Use the `setwd` R function to set the working directory to the local wrv-training repository specified above.

```{r}
setwd(path)
```

## Bugs

This website is a work in progress.
Please consider reporting bugs and asking for clarification on the
[Issues page](https://github.com/jfisher-usgs/wrv-training/issues).
Or you can just contact me via email at jfisher@usgs.gov
