---
title: "Overview"
---

Over the course of this 1-day training class you will learn to run
water use scenarios with the Wood River Valley (WRV) groundwater-flow model.

## Tentative Schedule

* 8:00--9:15  Welcome and introductions (Jim Bartoling, USGS)

* **8:15--9:15** Overview of WRV groundwater-flow-model publication, archive, and repository (Jason Fisher, USGS)

* 9:15--9:35 Break

* **9:30--10:30** Introduction to the R-programming language (Jason Fisher, USGS)

* **10:45--11:45** Running and analysing the archived [calibrated model](calib-model.html) (Jason Fisher, USGS)

* 11:45--12:45 Lunch

* 12:45--1:30 Running and analyzing the calibrated model (Jennifer Sukow, IDWR)

* 1:30--2:00 Running scenarios using the fully populated model, overview of example scenarios (Jennifer Sukow, IDWR)

* 2:00--2:15 Break

* 2:15--3:00 Scenerio 1: an example simulation of managed recharge (Jennifer Sukow, IDWR)

* **3:00--4:30** [Scenerio 2](scenario-2.html): an example simulation of idling irrigated lands (Jason Fisher, USGS)

## Files

The groundwater-flow model is described in the associated
[model report](http://dx.doi.org/10.3133/sir20165080 "USGS Scientific Investigations Report") and
[model archive](http://dx.doi.org/10.5066/F7C827DT "USGS Data Release").

The processing programs and unprocessed data sets for the groundwater-flow model
are bundled together in an R package named **wrv**.
The file repository for this package is located at https://github.com/USGS-R/wrv.
Included in the **wrv** package is [MODFLOW-USG](http://water.usgs.gov/ogw/mfusg/ "MODFLOW-USG") version 1.2,
a U.S. Geological Survey (USGS) computer code that solves the groundwater-flow equation.

Files used in this training class are available on a
[wrv-training repository](https://github.com/jfisher-usgs/wrv-training).
There will also be a USB-thumb drive distributed in class that contains the repository file structure,
and includes many other files that result from intermediate steps in data processing.
These additional files will be helpful if your computer is unable to execute some of processing steps
because of memory limitations or time constraints.

## Software

Installation instructions for most of the software used in this class is located at
https://jfisher-usgs.github.io/r-workshop/software.html.

A *beta* version of the R-package **wrv** also needs to be installed from its binary distribution file,
located on the USB drive ("wrv_1.0.3.zip").
This version of the package will soon be made available from the
**G**eological Survey **R** **A**rchive **N**etwork ([GRAN](https://owi.usgs.gov/R/gran.html)).
The new package version does not change model output or analysis results.
One major change is the removal of general-use functions from **wrv**.
These removed functions were bundled together in a new R package called **inlmisc**, available on
the **C**omprehensive **R** **A**rchive **N**etwork ([CRAN](https://cran.r-project.org/package=inlmisc)).

First remove any previous version of the **wrv** package you may have installed
by opening an R session and typing the following command:

```{r wrv_pkg_rm, eval=FALSE}
remove.packages("wrv")
```

To install the **inlmisc** package and its dependencies from CRAN use

```{r inlmisc_pkg_install, eval=FALSE}
install.packages("inlmisc", repos = "https://cloud.r-project.org/", dependencies = TRUE)
```

And to install the beta version of the **wrv** package use

```{r wrv_pkg_install, eval=FALSE}
install.packages(file.choose(), repos = NULL)
```
You will be prompted to select the location of the local ZIP file "wrv_1.0.3.zip".

## R programming

An introduction to the R-programming language is provided at
https://jfisher-usgs.github.io/r-workshop/index.html.

## Bugs

Please consider reporting bugs, requesting feature requests, and asking questions on the
[Issues page](https://github.com/jfisher-usgs/wrv-training/issues) for this website.
Or you can just contact me via email at jfisher@usgs.gov
