---
title: "Scenario"
---

```{r set_knitr_options, include=FALSE}
knitr::opts_chunk$set(eval=TRUE, dpi=100, fig.path="figures/")
```

These instructions describe how to build, run, and analyze an example water use scenario.
The scenario describes the effects of allowing an existing parcel of irrigated land to be reclassified as "non-irrigated";
that is, what happens when you voluntarily allow an irrigated parcel to go idle.
Model boundary conditions are modified in the archived calibrated model to best represent this scenario.

Startup an R session.
Set the working directory to the local wrv-training repository, that is the "wrv-training" folder,
by typing the following command in your R console and making sure to change the file path to the correct location:

```{r set_wd, eval=TRUE}
setwd("D:/WORK/Repos/wrv-training")
```

Load the **wrv**,
[**inlmisc**](https://CRAN.R-project.org/package=inlmisc),
[**raster**](https://CRAN.R-project.org/package=raster),
[**rgdal**](https://CRAN.R-project.org/package=rgdal), and
[**leaflet**](https://CRAN.R-project.org/package=leaflet)
packages into your current R environment, using

```{r load_pkgs, message=FALSE, results="hide", eval=TRUE}
packages <- c("wrv", "inlmisc", "raster", "rgdal", "leaflet")
lapply(packages, library, character.only = TRUE)
```

Draw a map the idled land parcel (about 600 acres, or better yet 2.4-square kilometers) and the 2002 land status
by pasting the following commands into your R console.

```{r leaflet_idle, fig.width=5.00, fig.height=5.00, eval=TRUE}
crs <- CRS("+init=epsg:4326")
irr <- spTransform(irr.lands[["2002"]], crs)
irr$Status <- factor(irr$Status, levels = c(levels(irr$Status), "idled"))
idx <- c(980, 981, 993, 1002, 1005, 1007, 1044, 1049, 1052,
         1059, 1061, 1065, 1066, 1067, 1073, 1074, 1076)
irr$Status[idx] <- "idled"
ll <- colMeans(coordinates(irr[idx, ]))
ext <- spTransform(alluvium.extent, crs)
irr <- crop(irr, extent(ext))
Pal <- colorFactor(c("#1B9E77", "#7570B3", "#D95F02"), irr$Status)
map <- leaflet()
map <- setView(map, lng = ll[1], lat = ll[2], zoom = 12)
url <- "http://basemap.nationalmap.gov/arcgis/services/USGSTopo/MapServer/WmsServer?"
opt <- WMSTileOptions(format = "image/png", transparent = TRUE)
map <- addWMSTiles(map, url, options = opt, layers = "0")
map <- addPolygons(map, data = irr, stroke = FALSE, fillOpacity = 0.6,
                   color = Pal(irr$Status))
map <- addPolylines(map, data = ext, weight = 3, color = "#000000")
map <- addLegend(map, pal = Pal, values = irr$Status, opacity = 0.6)
map
```

<p></p>

## File Changes

Changes to [unprocessed-data files](https://github.com/USGS-R/wrv/tree/master/inst/extdata) include:

* Irrigated lands ("extdata/irr/irr.lands.\<YYYY\>.zip")---changed parcel status to "non-irrigated".

* Evapotranspiration (ET)("extdata/et/et.\<YYYYMM\>.tif")---reduced ET within the parcel
  and set equal to the monthly precipitation during the irrigation season (April--October).

* Combined surface-water irrigation diversions ("extdata/div/comb.sw.irr.csv")--removed
  diversions associated with wells in parcel.

* Groundwater points of diversion ("extdata/div/pod.gw.csv")---removed diversions for wells in parcel.

* Well sites ("extdata/div/pod.wells.zip")---removed sites located in parcel.

Files that did not need to be modified for this scenario,
but may want to be considered for modification in similar types of scenarios include:

  * Surface water diversions ("div.sw.csv")---we assumed surface water associated
    with the idled parcel is now delivered to other "junior" users within the same canal service area.

  * Groundwater diversions ("div.gw.csv")---if there had been "measured" groundwater diversions
    associated with the parcel, those would need to be removed.

## Code Chunks

Most of the R code that we need for building, running, and analyzing the model scenario is contained within the appendix files.
Rather than rewrite all the relevant code here, we will instead use the `ReadCodeChunks` function in the **inlmisc** package
to read code chunks embedded within an appendix source file into our R session.
Read appendix C and D code chunks using the following commands:

```{r read_code_chunks, eval=TRUE}
file <- system.file("doc", "sir20165080_AppendixC.R", package = "wrv")
app.c.chunks <- ReadCodeChunks(file)
file <- system.file("doc", "sir20165080_AppendixD.R", package = "wrv")
app.d.chunks <- ReadCodeChunks(file)
```

Each of the relevant code chunks is labeled with a unique name.
For example, the following command retrieves the code chunk named "write_modflow_input" in appendix D:

```{r show_example_code_chunk, comment=NA, eval=TRUE}
chunk <- app.d.chunks[["write_modflow_input"]]
print(chunk)
```

## Create Modified Datasets

Every time we create a modified dataset it masks its **wrv**-package version.
The R script for creating the modified datasets is primarily based on code chunks in appendix C.
Run (that is, parse and evaluate) the code chunk named "setup" in appendix C using the following command:

```{r eval_setup, message=FALSE}
eval(parse(text = app.c.chunks[["setup"]]))
```

Specify the folder in your working directory that contains the modified unprocessed-data files ("extdata") and
place them in a temporary directory as uncompressed files using

```{r init_extdata, message=FALSE, results="hide"}
folder <- "extdata"
files <- list.files(folder, full.names = TRUE, recursive = TRUE)
dirs <- file.path(tempdir(), unique(dirname(files)))
sapply(dirs, function(i) dir.create(i, recursive = TRUE))
file.copy(files, file.path(tempdir(), files))
dir.in <- file.path(tempdir(), "extdata")
files <- list.files(dir.in, pattern = "*.zip$", full.names = TRUE, recursive = TRUE)
for (i in files) unzip(i, exdir = dirname(i))
```

Create a folder in your working directory where modified R-data files will be written.

```{r create_dir_out, message=FALSE}
dir.out <- "data"
dir.create(dir.out, showWarnings = FALSE)
```

Run the initial commands for specifying the unit conversions,
coordinate reference system (CRS), and common spatial/temporal grid.

```{r eval_unit_crs_spatial_temporal}
chunk.names <- c("unit_conversions", "crs", "spatial", "high_res_spatial", "temporal")
eval(parse(text = unlist(app.c.chunks[chunk.names])))
```

### Level 1 data

These are processing instructions for creating level 1 modified datasets.

#### Combined surface-water irrigation diversions (comb.sw.irr)

Read combined surface-water irrigation diversions data and write its external representation as an R object in "data/comb.sw.irr.rda".

```{r eval_comb_sw_irr}
eval(parse(text = app.c.chunks[["comb_sw_irr_1"]]))
```

#### Points of diversion for groundwater (pod.gw)

Read points of diversion for groundwater data and write its external representation as an R object in "data/pod.gw.csv.rda".

```{r eval_pod_gw}
eval(parse(text = app.c.chunks[["pod_gw_1"]]))
```

#### Well completions (pod.wells)

Read well completions data and write its external representation as an R object in "data/pod.wells.rda".

```{r eval_pod_wells}
eval(parse(text = unlist(app.c.chunks[c("pod_wells_1", "pod_wells_2")])))
```

#### Irrigation lands (irr.lands)

Read irrigated and semi-irrigated lands data and write its external representation as an R object in "data/irr.lands.rda".

```{r eval_irr_lands}
eval(parse(text = app.c.chunks[["irr_lands_1"]]))
```

#### Evapotranspiration (et)

Read evapotranspiration (ET) data and write its external representation as an R object in "data/et.rda".
Note that unprocessed ET raster files (".tif") are required for every month in the simulation,
not just months with modified values.

```{r eval_et_1}
eval(parse(text = unlist(app.c.chunks[c("et_1", "et_2")])))
```

### Level 2 data

These are processing instructions for creating level 2 modified datasets.
Recall that level 2 datasets are dependent on level 1 data.

#### Monthly irrigation entity components (entity.components)

Write an external representation of monthly irrigation entity components as an R object in "data/entitiy.components.rda".

```{r eval_entity_components}
eval(parse(text = unlist(app.c.chunks[paste0("entity_components_", 1:3)])))
```

#### Rasterized monthly irrigation entities (rs.entities)

Write an external representation of monthly irrigation entities as an R object in "data/rs.entities.rda"

```{r eval_rs_entities}
eval(parse(text = app.c.chunks[["rs_entities_1"]]))
```

#### Rasterized monthly recharge on non-irrigated lands (rs.rech.non.irr)

Write an external representation of monthly recharge on non-irrigated lands as an R object in "data/rs.rech.non.irr.rda".

```{r eval_rs_rech_non_irr}
eval(parse(text = app.c.chunks[["rs_rech_non_irr_1"]]))
```

## Model Processing

Let's start model processing with a clean R environment.
Remove all objects created during the previous instructions and reload our modified datasets.

```{r clean_env, results="hide", eval=TRUE}
rm(list = ls()[!ls() %in% "app.d.chunks"])
files <- list.files("data", pattern = "*.rda$", full.names = TRUE)
lapply(files, load, envir = .GlobalEnv)
```

### Copy calibrated files

Copy all the [calibrated model](calib-model.html) files from the "archive" folder
to a new folder named "model" in our working directory.

```{r copy_archive_to_model, results="hide"}
dir.create("model", showWarnings = FALSE)
files <- list.files("archive", full.names = TRUE)
file.copy(files, "model", overwrite = TRUE)
```

Specify the identifier for the model run.

```{r set_id, results="hide", eval=TRUE}
id <- "wrv_mfusg"
```

Load R objects from the "model/model.rda" file into your R environment, these objects include
`d.in.mv.ave`, `misc`, `reduction`, `rs`, `ss.stress.periods`, `tr.stress.periods`, and `trib`.
A description of each R object is given in the help documentation of the
`UpdateWaterBudget` function in the **wrv** package.

```{r load_model, results="hide", eval=TRUE}
load("model/model.rda")
```

The `rs` object in the appendix D code is named `rs.model`; use the following command to address this issue:

```{r assign_rs, eval=TRUE}
rs.model <- rs
```

### Recalculate Specified Flows

Recalculate the specified-flow boundary conditions using our [modified datasets](scenario.html#create_datasets)---this
results in a new MODFLOW Well file ("model/wrv_mfusg.wel") and quality assurance tables ("model/qa-\*.csv").
Note that all modified datasets need to be passed as arguments in the `UpdateWaterBudget` function.

```{r update_water_budget, results="hide"}
UpdateWaterBudget("model", id, qa.tables = "english", pod.wells = pod.wells,
                  comb.sw.irr = comb.sw.irr, et = et,  pod.gw = pod.gw,
                  entity.components = entity.components, rs.entities = rs.entities)
```

### Run MODFLOW

Run [MODFLOW-USG](http://water.usgs.gov/ogw/mfusg/) using model scenario conditions.

```{r run_modflow}
wd <- setwd("model")
system2(file.path(getwd(), "RunModflow.bat"), stdout = FALSE, stderr = FALSE)
setwd(wd)
```

### Read water budgets

Read water-budget output for the archived calibrated simulation,
located in the "archive" folder, and save the results for later use.
To do so, run the code chunks named "read_budget_1" and "read_budget_2"
in appendix D using the following commands:

```{r eval_budget_archive, message=FALSE, warning=FALSE, eval=TRUE}
dir.run <- "archive"  # code below is dependent on this object
eval(parse(text = unlist(app.d.chunks[paste0("read_budget_", 1:2)])))
```

And save the average volumetric flow rates to an R object named `budget1` using

```{r save_budget_archive, eval=TRUE}
budget1 <- c("Water-table recharge"             = mean(d.rech$flow.in),
             "Streamflow losses"                = mean(d.river$flow.in),
             "Tributary basin underflow"        = mean(d.trib$flow),
             "Water-table discharge"            = mean(d.rech$flow.out),
             "Streamflow gains"                 = mean(d.river$flow.out),
             "Production well pumping"          = mean(d.well$flow),
             "Stanton Crossing outlet boundary" = mean(d.drain.1$flow),
             "Silver Creek outlet boundary"     = mean(d.drain.2$flow))
budget1 <- setNames(as.integer(abs(budget1)) * 0.296106669, names(budget1))
```

Next, read the water-budget output for the scenario simulation, located in the "model" folder.

```{r eval_budget_scenario, message=FALSE, warning=FALSE, eval=TRUE}
dir.run <- "model"
eval(parse(text = unlist(app.d.chunks[paste0("read_budget_", 1:2)])))
```

And save the average volumetric flow rates to an R object named `budget2` using

```{r save_budget_scenario, eval=TRUE}
budget2 <- c("Water-table recharge"             = mean(d.rech$flow.in),
             "Streamflow losses"                = mean(d.river$flow.in),
             "Tributary basin underflow"        = mean(d.trib$flow),
             "Water-table discharge"            = mean(d.rech$flow.out),
             "Streamflow gains"                 = mean(d.river$flow.out),
             "Production well pumping"          = mean(d.well$flow),
             "Stanton Crossing outlet boundary" = mean(d.drain.1$flow),
             "Silver Creek outlet boundary"     = mean(d.drain.2$flow))
budget2 <- setNames(as.integer(abs(budget2)) * 0.296106669, names(budget2))
```

### Compare water budgets

A comparison between water budgets is made by constructing the following table.

**Table:** Water budgets for modified and archived simulation, specified as volumetric flow rates average over 1998--2010.
[**Inflow:** water entering the aquifer system.
**Outflow:** water leaving the aquifer system.
**Component:** a water budget component in the groundwater-flow model.
**Archived Rate:** is the mean volumetric flow rate for the archived simulation.
**Modified Rate**: is the mean volumetric flow rate for the modified simulation.
**Difference:** is the archived rate subtracted from the modified rate.
**Abbreviations:** acre-ft/yr, acre-feet per year]

```{r table_budget, eval=TRUE}
d <- data.frame(rate1 = budget1, rate2 = budget2)
d$direction <- c("**Inflow**", "", "", "**Outflow**", "", "", "", "")
d$component <- names(budget1)
d$diff <- budget2 - budget1
d$pchange <- ((budget2 - budget1) / budget1) * 100
d <- rbind(d[1:3, ], NA, d[4:8, ], NA, NA)
d$component[nrow(d)] <- "**Inflow - Outflow**"
d$rate1[nrow(d)] <- sum(budget1[1:3]) - sum(budget1[5:8])
d$rate2[nrow(d)] <- sum(budget2[1:3]) - sum(budget2[5:8])
rownames(d) <- NULL
d <- d[, c("direction", "component", "rate1", "rate2", "diff", "pchange")]
colnames(d) <- c("", "Component", "Archived rate<br>(acre-ft/yr)",
                 "Scenario rate<br>(acre-ft/yr)", "Difference<br>(arce-ft/yr)",
                 "Percent<br>change")
d[, 3] <- formatC(d[, 3], format = "f", digits = 0, big.mark = ",")
d[, 4] <- formatC(d[, 4], format = "f", digits = 0, big.mark = ",")
d[, 5] <- formatC(d[, 5], format = "f", digits = 0, big.mark = ",")
d[, 6] <- formatC(d[, 6], format = "f", digits = 1, big.mark = ",")
d[, 1:2][is.na(d[, 1:2])]  <- ""
d[, 3:6][d[, 3:6] == "NA"] <- ""
knitr::kable(d, format = "markdown", padding = 0, booktabs = TRUE, digits = 0,
             row.names = FALSE, col.names = colnames(d), align = "llrrrr")
```

### Read hydraulic heads

Read simulated hydraulic heads for archived simulation, located in "archive" folder, and save results.

Read the simulated hydraulic head (head) output for the archived calibrated model,
located in the "archive" folder, and save the results for later use.
To do so, run the code chunk named "read_head" in appendix D using the following commands:

```{r eval_head_archive, message=FALSE, warning=FALSE, eval=TRUE}
dir.run <- "archive"
eval(parse(text = app.d.chunks[["read_head"]]))
head1 <- rs.heads.lay1
```

Next, read the head output for the scenario simulation, located in the "model" folder.

```{r eval_head_scenario, message=FALSE, warning=FALSE, eval=TRUE}
dir.run <- "model"
eval(parse(text = app.d.chunks[["read_head"]]))
head2 <- rs.heads.lay1
```

### Compare head distributions

Draw a map the simulated water table difference,
defined as the archived water table subtracted from the modified water table, during August 16, 2007.
Add an observation point located at a longitude and latitude of -114.222664 and  43.383047 decimal degrees, respectively.

```{r leaflet_head, fig.width=5.00, fig.height=5.00, eval=TRUE}
ext <- spTransform(alluvium.extent, CRS("+init=epsg:4326"))
r <- head2[["2007-08-16"]] - head1[["2007-08-16"]]
r[] <- round(r[] * 3.28084, digits = 3)
r <- projectRaster(r, crs = CRS("+init=epsg:4326"), method = "ngb")
Pal <- colorNumeric("Spectral", r[], na.color = "transparent")
map <- leaflet()
ll  <- coordinates(ext)
map <- setView(map, lng = ll[1], lat = ll[2], zoom = 11)
url <- "http://basemap.nationalmap.gov/arcgis/services/USGSTopo/MapServer/WmsServer?"
opt <- WMSTileOptions(format = "image/png", transparent = TRUE)
map <- addWMSTiles(map, url, options = opt, layers = "0")
map <- addRasterImage(map, r, colors = Pal, opacity = 0.8)
map <- addPolylines(map, data = ext, weight = 3, color = "#000000")
map <- addLegend(map, pal = Pal, values = r[], opacity = 0.8,
                 title = "Head Diff.<br/>(feet)")
ll <- c(-114.222664, 43.383047)
txt <- sprintf("<b>Longitude:</b> %s<br/><b>Latitude:</b> %s", ll[1], ll[2])
icon <- icons(iconUrl = "markers/marker-blue.png",
              iconWidth = 34, iconHeight = 34, iconAnchorX = 17, iconAnchorY = 34)
map <- addMarkers(map, lng = ll[1], lat = ll[2], popup = txt, icon = icon)
map
```

<p></p>

Finally, at the observation point, plot the head difference in model layer 1 over time.

```{r graph_head_diff, fig.width=7.16, fig.height=3.50, eval=TRUE}
pnt <- data.frame(lon = ll[1], lat = ll[2])
coordinates(pnt) <- c("lon", "lat")
proj4string(pnt) <- CRS("+init=epsg:4326")
pnt <- spTransform(pnt, crs(hill.shading))
d <- t(extract(head2, pnt) - extract(head1, pnt))
d <- data.frame(Date = as.Date(rownames(d)), difference = d)
ylab <- paste("Hydraulic head difference, in", c("meters", "feet"))
PlotGraph(d, ylab = ylab, col = "#025D8C", conversion.factor = 3.28084,
          center.date.labels = TRUE)
```
